### 2022-11-10 chapter1.tex
Removed line numbers and added asterisk to corresponding author
Corrected some minor typos from Colleen and Julia

\usepackage[margin=1in]{geometry}
\usepackage[right]{lineno}  % line numbers
%\usepackage{setspace}  % double spacing
%\doublespacing
\usepackage[skip=10pt]{parskip}

@@

\begin{document}
\linenumbers

% Title must be 250 characters or less
\begin{flushleft}
{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
\newline
\\
Marc Singleton\textsuperscript{1},
Michael Eisen\textsuperscript{1,2,3}
\\
\bigskip
\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
\\
\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
\\
\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
\\
\bigskip

@@

\section*{Introduction}
Comparative genomics is powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflects variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved and equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig. \ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 \ref{sfig:hmm_trim}.

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller \cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon-sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@

\nolinenumbers

>>>>

\usepackage[margin=1in]{geometry}
%\usepackage{setspace}  % double spacing
%\doublespacing
\usepackage[skip=10pt]{parskip}

@@

\begin{document}

% Title must be 250 characters or less
\begin{flushleft}
{\Large\textbf{Leveraging genomic redundancy to improve inference and alignment of orthologous proteins}}
\newline
\\
Marc Singleton\textsuperscript{1},
Michael Eisen\textsuperscript{1,2,3,*}
\\
\bigskip
\textbf{1} Biophysics Graduate Group, UC Berkeley, Berkeley, CA, USA
\\
\textbf{2} Department of Molecular and Cell Biology, UC Berkeley, Berkeley, CA, USA
\\
\textbf{3} Howard Hughes Medical Institute, UC Berkeley, Berkeley, CA, USA
\\
\bigskip

@@

\section*{Introduction}
Comparative genomics is a powerful tool for yielding insights into evolutionary relationships, molecular function, and the forces that drive gene, genome, and population evolution. These methods often rely on the identification of homologous sequences or homologs, that is sequences with common ancestry, since this ensures that differences between sequences reflect variations in evolution from a common point of divergence. In a sense, each branch that descends from a single ancestral population is a different ``experiment'' by nature, as each is free to explore the evolutionary landscape independent of the others. However, many analyses impose the additional condition that the sequences have diverged through speciation events (orthology) rather than duplications (paralogy) or other mechanisms such as horizontal gene transfer \cite{Fitch1970}. The underlying assumption is ``orthologs'' have conserved equivalent functions whereas ``paralogs'', by virtue of their redundancy, are more likely to diverge \cite{Ohno1970, Nowak1997, Altenhoff2012, Pegueroles2013, Soria2014}.\footnote{For multiple sequences, orthology is usually defined relative to their most recent common ancestor. This technically includes sequences which split by duplication after this point (``in-paralogs''), but excludes sequences which split by duplication before (``out-paralogs'') \cite{Remm2001}. Many current orthology inference pipelines explicitly incorporate steps to detect in-paralogs. However, the resulting orthologs groups can easily be restricted to those without in-paralogs (``single copy orthologs'') for analyses where an assumption of conserved function is necessary.} This relationship between orthology and function is an essential component of modern biological research since it permits the transfer of annotations between biological systems using sequence similarity alone.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states.” Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. This phylo-HMM was composed of four hidden states, {1A, 1B, 2, 3}, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment (Fig. \ref{fig:hmm}). Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. Fewer than 10\% of alignments were trimmed, but when regions were trimmed, they were largely inferred as state 3 (Fig. \ref{sfig:hmm_trim}).

@@

The second major innovation in this work is its method for the refinement and trimming of alignments. In refinement, a profile HMM is constructed from the well-aligned columns of an initial alignment, and the original sequences are aligned to this profile. This process usually corrects errors caused by the presence of long poorly supported segments. In trimming, a phylo-HMM is used to remove regions and segments which are poorly supported by the phylogenetic consensus. While the combination of these two steps yielded high-quality alignments that are suitable for further analyses, they are an \textit{ad hoc} fix for underlying issues with the gene models and alignment algorithms. The most principled solution is to optimize or supplement the gene models using the initial alignments generated by the orthology inference pipeline, which is possible with tools such as OMGene or OrthoFiller \cite{Dunne2018, Dunne2017}. However, if preserving the original annotations is desired or necessary, the remaining possibility is to correct the alignments as we have done here. In fact, the errors we sought to address broadly stem from shortcomings of current alignment algorithms rather than errors in the sequences themselves. Though the scoring functions of modern multiple sequence alignment algorithms are complex, they are generally derived from models that penalize gaps with a linear or affine cost. As a result, they often interlace gaps with short, aligned segments rather than a single long gap. However, when the sequences are different isoforms of a single gene, their alignment will necessarily contain contiguous exon sized gaps. The same is true when aligning isoforms of diverged orthologs, though the relationship between their exons may be complex.

@@
